Differential effects of knapping skill acquisition on the cultural reproduction of Late Acheulean handaxe morphology: Archaeological and experimental insights

### 2023-02-01 manuscript/manuscript.Rmd
integrating more of Dietz's comments.

---
title: "Differential effects of knapping skill on the cultural reproduction of Late Acheulean handaxe morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Rock Art Research Institute, School of Geography, Archaeology, and Environmental Studies, University of the Witwatersrand, Johannesburg, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapper skill levels and mental templates. Integrating these two lines of literature into a broader theoretical framework of cultural reproduction, here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that knapper skill levels and mental templates have a relatively clear manifestation in different aspects of handaxe morphology. The former relates to cross-sectional thinning (PC1), while the latter refers to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training on the differential cultural reproduction of these two aspects using the data from a 90-hour-long knapping skill acquisition experiment. We found that the desired shape of a handaxe can be relatively quickly picked up by novices, while reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural transmission"
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

In its classical definition, the term mental template indicates that the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45]. This concept lies at the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework, which also great overlaps with some core assumptions of the more modern approach of cultural transmission theory [@eerkens2005; @eerkens2007]. Early researchers, whether explicitly or implicitly, often endorsed this conceptual framework and actively applied it in the typological analysis of handaxes at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004]. Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as a major technological innovation of the Acheulean compared with the Oldowan [@isaac1986].

For a decade or so, this concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], b) ignoring other competing factors such as raw material constraints [@white1995], and c) being constrained by the basic fracture mechanics and design space of bifacial technology [@moore2011; @moore2016]. A more recent approach has been to identify morphological "design imperatives" derived from utilitarian and ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018]. The major difference between the concepts of design imperatives and mental templates lies in the fact that the former does not necessarily require the presence of explicit internal representations of form, where the shape of handaxes can instead emerge "through the coalescence of ergonomic needs in the manipulation of large cutting tools [@wynn2021: 185]." Following this discussion, Kuhn [-@kuhnEvolutionPaleolithicTechnologies2020: 168-170] developed a complimentary framework by explicitly identifying how different factors constrain the morphology of the design target, such as production constraint (raw materials) and functional constraint (mechanical and symbolic factors).

@@

The cultural reproduction, or cultural transmission as described in standard cultural evolutionary literature [@eerkens2005; @eerkens2007], of mental templates and skill levels makes them reach beyond individual-level practice and form a repetitive pattern that can be identified in archaeological records. Nonetheless, the abstract shape of handaxe as a mental template that is often pulled away from its original substrate has been frequently treated as the main research subject of cultural transmission experiments [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015], while how knapping skill as another source of variation is reproduced during the learning process and how it moderates the material manifestation of mental templates has been rarely discussed. The ignorance of the latter becomes one of motivations behind our terminological choice of "reproduction" over "transmission", where the former implies more than just the copying of an static image with information loss [@liu2022; @stout2021a]. This reframing essentially echoes the stance of extended evolutionary synthesis (EES) on inclusive inheritance that phenotypes are not inherited but reconstructed in development [@laland2015: 5], which has also received more attention recently in the domain of cultural evolution [@charbonneau2022; @strachan2021].

Centering around the concept of cultural reproduction, we aim to explore the possibility of dissecting the interaction of skill level and mental template through a comparative study of an archaeological handaxe assemblage known for its remarkable high skill level, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers. We generated the novice handaxe collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest. As such, our theory-driven data-informed project has the following two interconnected research questions: 1) What can the deep structure revealed through the multivariable analysis of handaxe morphometric data inform us on the material manifestation of knapping skills and mental templates? Our presumption here is that the morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect skill level differences, and all three group should show a similar mental template since this is a common target. 2) Does training has a differential effect in terms of the reproduction of knapping skill level and mental templates among novices? Our expectation is that throughout the training the novice samples should become more similar to expert samples in both skill level and mental template, but the acquisition of the former aspect will be more challenging and thereby slower than the latter aspect. This hypothesis is informed by the previous study of Pargeter et al. [-@pargeter2020] showing that in handaxe manufacture novices' predictions of the contour of flakes to be removed are highly similar to those of expert knappers, while novices do not have the right forces and accuracy to successfully remove their target flakes to produce a nice handaxe.

@@

Our study suggests that skill can differentially affect the expression and of different aspects of artifact mental templates, potentially biasing processes of cultural reproduction. In the case of handaxe morphology, we found that skill is more highly constraining of cross-sectional thinning (PC1) than it is of handaxe elongation and pointedness (PC2). This is in accordance with the existing literature on handaxe knapping skill [@caruana2020; @stout2014; @callahan1979], and supports the use of cross-sectional thinning as a robust indicator of skill level at Boxgrove. It also suggests that cultural evolutionary approaches to handaxe morphology should consider technological choices about investment in skill acquisition [@pargeter2019] as a directional influence alongside random copy error [@eerkens2005] as sources of variation. In contrast, we found morphological targets not requiring cross-sectional thinning (elongation and pointedness (PC2)) to be less constrained by skill. These aspects of morphology might thus provide a clearer signal of "arbitrary" cultural variation and accumulating copy error. Notably, Boxgrove handaxes are highly constrained along PC2 compared to our experimental samples, in keeping with prior arguments that production at this site adhered to of a well-defined mental template [@garcía-medrano2019; @shipton2020].

Thinning is regarded as a technique requiring a high knapping skill level because it requires one to carefully detach flakes in an invasive manner while not breaking the handaxe into several pieces, serving the purpose of achieving the desired convexity and/or volume. This procedure involves precise control of striking forces, strategic choice of platform external angle, and attentive preparation of bifacial intersection plane, all of which were part of our experimental training program [@callahan1979; @caruana2022; @stout2014; @pargeter2020; @shipton2013]. Experimental studies have also shown that the thinning stage of handaxe produce often involves the use of soft hammers, which is also supported by indirect archaeological evidence of flake attributes from Boxgrove [@roberts2009; @roberts1998: 384-394], although the validity of differentiating purcussor types (hard hammerstone, soft hammerstone, and antler hammer) based on flake attributes has been challenged by other experimental studies[@driscoll2014]. It should be noted that both our experts and novices frequently used soft hammers in the production of experimental assemblages. In the skill acquisition experiments, novice knappers were explicitly taught to switch to the soft hammer for thinning purposes, but some of them did not follow the instruction during the assessment. On the other hand, it has also been shown that hard hammers can also be used to achieve similar thinning results [@pelcin1997; @bradley1986], and the replicas produced by Bruce Bradley in our expert reference collection did not involve the use of soft hammers.

@@

# Conclusions

Regarding the two research questions we proposed in the beginning, our case study suggested that 1) To some extent we can delineate the effects of knapping skill and mental template through the handaxe morphometric data, where the former is closely associated with cross-sectional thinning while the latter is mainly expressed in elongation and pointedness due to the constraint of the former; 2) On average training has an immediate effect of making novices to better understand the shared design targets, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove tool makers, which supports our differential cultural reproduction hypothesis. At a larger theoretical level it questions the distinction between social learning of design targets vs. individual learning of the skills needed to achieve them. Traditionally archaeological experiments speaking to the literature of cultural evolution tend to use handaxe as a model artifact and focus on how copying errors emerge during the transmission of a fixed and static target using transmission chain design and alternative raw materials such as foam [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015]. This line of inquiry is generally characterized by high internal validity (causal mechanisms) but low external validity (generalizability to archaeological data). In contrast, our study unpacks the differential reproductions of two major sources of variation and reveals how the development of motor skill during learning is constraining the achievement of the socially learnt design target, through an actualistic experimental setting featuring a higher degree of external validity [@liu2022]. In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes in the proper developmental context [@högberg2018; @nowell2021; @lew-levy2020] as well as their implications for the biological and cultural evolution of the hominin lineages.

>>>>

---
title: "Differential effects of knapping skill acquisition on the cultural reproduction of Late Acheulean handaxe morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Rock Art Research Institute, School of Geography, Archaeology, and Environmental Studies, University of the Witwatersrand, Johannesburg, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapper skill levels and mental templates. Integrating these two lines of literature into a broader theoretical framework of cultural reproduction, here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that knapping skill acquisition has a differnetial effect in the cultural reproduction of different aspects of handaxe morphology. More specifically, compared with elongation and pointedness (PC2), cross-sectional thinning (PC1) is more constrained by knapping skill. It also confirmed that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running (90-hour) training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural transmission"
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

In its classical definition, the term mental template indicates that the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45]. This concept lies at the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework. Early researchers, whether explicitly or implicitly, often endorsed this conceptual framework and actively applied it in the typological analysis of handaxes at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004]. Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as a major technological innovation of the Acheulean compared with the Oldowan [@isaac1986]. Importantly, this conception of a mental template as an idea or image transmitted between minds also echoes core assumptions of the more modern approach of cultural transmission theory [@eerkens2005; @eerkens2007].

For a decade or so, the mental template concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], b) ignoring other competing factors such as raw material constraints [@white1995], and c) being constrained by the basic fracture mechanics and design space of bifacial technology [@moore2011; @moore2016]. A more recent approach has been to identify morphological "design imperatives" derived from utilitarian and ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018]. The major difference between the concepts of design imperatives and mental templates lies in the fact that the former does not necessarily require the presence of explicit internal representations of form, where the shape of handaxes can instead emerge "through the coalescence of ergonomic needs in the manipulation of large cutting tools [@wynn2021: 185]." Following this discussion, Kuhn [-@kuhnEvolutionPaleolithicTechnologies2020: 168-170] developed a complimentary framework by explicitly identifying how different factors constrain the morphology of the design target, such as production constraint (raw materials) and functional constraint (mechanical and symbolic factors).

@@

The cultural reproduction, or transmission as it is commonly termed in the cultural evolutionary literature [@eerkens2005; @eerkens2007], of mental templates and production skills makes them reach beyond individual-level practice and form a repetitive pattern that can be identified in archaeological records. Nonetheless, the abstract shape of handaxe as a mental template that is often pulled away from its original substrate has been frequently treated as the main research subject of cultural transmission experiments [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015], while how knapping skill as another source of variation is reproduced during the learning process and how it moderates the material manifestation of mental templates has been rarely discussed. The ignorance of the latter becomes one of motivations behind our terminological choice of "reproduction" over "transmission", where the former implies more than just the copying of an static image with information loss [@liu2022; @stout2021a]. This reframing essentially echoes the stance of extended evolutionary synthesis (EES) on inclusive inheritance that phenotypes are not inherited but reconstructed in development [@laland2015: 5], which has also received more attention recently in the domain of cultural evolution [@charbonneau2022; @strachan2021].

Centering around the concept of cultural reproduction, we aim to explore the possibility of dissecting the interaction of skill level and mental template through a comparative study of an archaeological handaxe assemblage known for its remarkable high skill level, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers. We generated the novice handaxe collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest. As such, our theory-driven data-informed project aims to examine the following research questions: 1) Do the processes of skill learning in a lithic medium exert any biases on the cultural reproduction of artifact morphology? To answer this question, we first applied multivariate analysis of handaxe morphometric data to identify different components of morphological variation. Our study has the following three assumptions: 1) Boxgrove, expert, and novice assemblages should have a common design target (mental template) as Boxgrove is designated as the model during the training; 2) The morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect skill level differences; 3) Throughout the training the novice samples should become more similar to expert samples in both skill level and mental template, but the acquisition of the former aspect will be more challenging and thereby constraining the latter aspect. The third assumption here is oarticularly informed by the previous study of Pargeter et al. [-@pargeter2020] showing that in handaxe manufacture novices' predictions of the contour of flakes to be removed are highly similar to those of expert knappers, while novices do not have the right forces and accuracy to successfully remove their target flakes to produce a nice handaxe.

@@

Our study suggests that skill can differentially affect the expression of different aspects of artifact mental templates, potentially biasing processes of cultural reproduction. In the case of handaxe morphology, we found that skill is more highly constraining of cross-sectional thinning (PC1) than it is of handaxe elongation and pointedness (PC2). This is in accordance with the existing literature on handaxe knapping skill [@caruana2020; @stout2014; @callahan1979], and supports the use of cross-sectional thinning as a robust indicator of skill at Boxgrove, with potential implications for the cognitive demands and social contexts supporting learning [@pargeter2019; @stout2014]. It further suggests that cultural evolutionary approaches to handaxe morphology should consider technological choices about investments in skill acquisition [@pargeter2019] as a directional influence alongside random copy error [@eerkens2005] as sources of variation. In contrast, we found morphological targets not requiring cross-sectional thinning (elongation and pointedness (PC2)) to be less constrained by skill. These aspects of morphology might thus provide a clearer signal of "arbitrary" cultural variation and accumulating copy error. Notably, Boxgrove handaxes are highly constrained along PC2 compared to our experimental samples, in keeping with prior arguments that production at this site adhered to of a well-defined mental template [@garcía-medrano2019; @shipton2020].

Thinning is regarded as a technique requiring a high knapping skill level because it requires one to carefully detach flakes in an invasive manner while not breaking the handaxe into several pieces, serving the purpose of achieving the desired convexity and/or volume. This procedure involves precise control of striking forces, strategic choice of platform external angle, and attentive preparation of bifacial intersection plane, all of which were part of our experimental training program [@callahan1979; @caruana2022; @stout2014; @pargeter2020; @shipton2013]. Experimental studies have also shown that the thinning stage of handaxe produce often involves the use of soft hammers, which is also supported by direct [@stout2014; @bello2016] and indirect [@roberts2009; @roberts1998: 384-394] archaeological evidence from Boxgrove, although the validity of differentiating purcussor types (hard hammerstone, soft hammerstone, and antler hammer) based on flake attributes has been challenged by other experimental studies[@driscoll2014]. It should be noted that both our experts and novices frequently used soft hammers in the production of experimental assemblages. In the skill acquisition experiments, novice knappers were explicitly taught to switch to the soft hammer for thinning purposes, although some of them did not follow the instruction during the assessment. On the other hand, it has also been shown that hard hammers can also be used to achieve similar thinning results [@pelcin1997; @bradley1986], and the replicas produced by Bruce Bradley in our expert reference collection did not involve the use of soft hammers.

@@

# Conclusion

Regarding the research question we proposed in the beginning, our case study suggested that skill acquisition has a differential effect in the cultural reproduction of different aspects of mental templates, where cross-sectional thinning (PC1) is more constrained by knapping skill while elongation and pointedness (PC2) is less so. At a larger theoretical level it questions the distinction between social learning of design targets vs. individual learning of the skills needed to achieve them. Traditionally archaeological experiments speaking to the literature of cultural evolution tend to use handaxe as a model artifact and focus on how copying errors emerge during the transmission of a fixed and static target using transmission chain design and alternative raw materials such as foam [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015]. This line of inquiry is generally characterized by high internal validity (causal mechanisms) but low external validity (generalizability to archaeological data). In contrast, our study unpacks the differential reproductions of two major sources of variation and reveals how the development of motor skill during learning is constraining the achievement of the socially learnt design target, through an actualistic experimental setting featuring a higher degree of external validity [@liu2022]. In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes in the proper developmental context [@högberg2018; @nowell2021; @lew-levy2020] as well as their implications for the biological and cultural evolution of the hominin lineages.
